Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-04-01 ms/spaceness.tex
ms updates

%Simulating Isolation by Distance in Continuous Space
%Space is the Place: How Dispersal and Competition Shape Genetic Variation in Continuous Space
%Impacts of Continuous Spatial Structure on Analyses of Population Genetic Data
%Impacts of Isolation by Distance on Demographic Modeling and Association Studies
%Impacts of Dispersal and Sampling on Estimates of Genetic Variation in Continuous Space
%Space is the Place: How Dispersal and Sampling Shape Estimates of Genetic Variation in Continuous Space
\title{Space is the Place: How Dispersal and Sampling Shape Analyses of Genetic Variation in Continuous Space}

@@

\affil[$\ast$]{University of Oregon Dept. Biology, Institute for Ecology Evolution}
\keywords{Space; Population Structure; Demography; Haplotype block sharing}

@@

\begin{abstract}
Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations exchanging migrants. Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and association studies. Low dispersal slows the geographic spread of ancestry and increases branch lengths corresponding to mid-frequency alleles because of slow coalescence among distant groups of individuals. As a result demographic inference from the site frequency spectrum is biased towards inferring a turbulent past demographic history and a recent decline in populations when neighborhood sizes are below 20. For many common population genetic summary statistics the variation associated with different sampling strategies is larger than that observed across the range of dispersal distances we simulated, pointing to the need to carefully account for the spatial distribution of sampling when interpreting results. Last we show that spatially correlated environments cause genome-wide signals of association with purely environmental phenotypes in linear-regression GWAS, and that this bias is not fully controlled by regressing out principal components positions during the analysis.  
\end{abstract}

@@

\section{Introduction}
The inescapable reality that biological organisms live, move, and reproduce in continuous spatial landscapes has been all but ignored in population genetic models. Indeed a near universal rule of reproduction is that individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances. This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but our mathematical models to describe it are poor approximations of the underlying process. As the availability of population-level, whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating biologically realistic demographic and spatial processes are needed to accurately describe continuous spatial structure and control for its impacts on analyses of selection and demography. This situation is particularly clear in human population genetics, where individual ancestries are often interpreted as mixtures of discrete populations in part because of modeling assumptions underlying discrete clustering algorithms and association analyses are often strongly dependent on statistical corrections for population stratification \citep{Berg2018,Bulik-Sullivan2015,Young2018,Heckerman2016,Kang2010,Sohail2018}.

The best-studied approaches to population genetics in continuous space were developed by \citep{Wright1943} and \citep{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. A key finding of Wright's model is that many important aspects of continuous populations can be described in terms of "neighborhood size" -- the number of potential mates for an individual in a given generation, defined as $4\pi\sigma^2d$, where $\sigma$ is the average dispersal distance and $d$ is population density. \cite{Maruyama1972} found that the rate of decline in genetic diversity in a 2-dimensional continuous population approaches the random mating expectation when $D\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

Though some aspects of continuous populations are fairly well described by the Wright-Malécot models, \cite{Felsenstein1975} showed that the assumptions of independent dispersal and Poisson distributed offspring that are the basis of these models are incompatible. Over time, a population meeting them will clump into a small number of geographic clusters occupying only a part of the available range. Although real populations are often clumped on landscapes due to factors like varying habitat quality and competition among species, the Wright and Malécot models produce much more extreme clumping than is observed in practice and fail to account for the density-dependent declines in population growth rate that are widely observed in real populations (CITE).

One method for modeling continuous populations is then to assume the existence of a grid of discrete randomly-mating populations connected by migration, which prevents clustering by forcing all regions to be occupied in every generation. Among many other important results drawn from this class of "lattice" or "stepping stone" models, \cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{st}$) against the logarithm of spatial distance is an estimate of neighborhood size. Though good approximations of continuous structure given high dispersal (Barton(?)), these models are not truly continuous, force a uniform realized population density across landscapes, and limit investigation of spatial structure below the level of the deme. An alternative method is to model the geographic spread of ancestry backwards in time through a diffusion approximation -- an approach that has recently made significant progress in modeling both dispersal and demographic parameters \citep{Barton2010,Kelleher2014,Ringbauer2017,Ringbauer2018}. However these models currently lack a well-described forward-time analog that produces biologically realistic spatial patterns and are subject to the same difficulty in incorporating selection and other biological processes as all other reverse time methods (CITE??).

\ak{Andy-- I think we could move some of the text from the following paragraph to methods}
We took a direct approach to the clustering problem of classical forward-time models by incorporating density dependence into the model. By scaling the probability of survival in each timestep to local population density we shift reproductive output towards regions of low-density and prevent populations from clustering. A similar approach was taken previously by \citep{Doebeli2003} who used an individual based model with continuous space and density dependent fitness to study the probability of speciation along continuous environmental gradients. However to our knowledge all previous implementations of continuous space models have focused on a small number of genetic loci as the unit of analysis, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms.

Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness. The model scales to chromosome-scale alignments across tens of thousands of individuals, and outputs the full genealogy and recombination history of all final-generation individuals. We use this simulation to describe how sampling strategy interacts with isolation by distance to cause systematic variation in population genetic summary statistics typically analyzed under discrete population models. We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum; finding that inflation of branch lengths for mid-frequency alleles associated with slow coalescence of distant individuals can bias analyses towards inferring turbulent demographic histories and recent declines in population size.

Last, we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS). Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations \citep{Price2006}. Because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within continuous populations can also deflate $p$ values in GWAS of quantitative traits \citep{Mathieson2012}. These issues are particularly prominent for polygenic traits, because the deflated $p$ values expected from polygenic phenotype correlations are superficially similar to those expected from population structure \citep{Bulik-Sullivan2015,Young2018}. However most studies examining the effects of stratification on GWAS have focused on simulating phenotypes for real genotype data \citep{Price2006,Bulik-Sullivan2015} or generating simulated genotypes by sampling from static allele frequency distributions \citep{Bulik-Sullivan2015,Heckerman2016}. These approaches address application to human data but are not based on a true generative model and so provide limited insight into the processes generating stratification bias in association analyses. Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e. the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods of stratification bias corrections in GWAS are subject to bias.

%[move some of this to methods, rewrite to avoid repitition w above]  A common method of correcting for population structure in GWAS is to run a principal components analysis on the genotype matrix and include PC coordinates as covariates in a linear regression for each SNP \citep{Price2006}. Mixed model GWAS employs a different statistical framework but takes a similar approach by incorporating a kinship matrix describing pairwise genetic relatedness among samples as a random effect in a linear mixed model \citep{Yu2006,Kang2008}. In both approaches the key question is whether or not the ancestry components of the model adequately describe "background" levels of genetic variation in the sample relative to any confounding effects of ancestry and phenotype. Recent studies in European and British populations \citep{Kerminen2018,Berg2018,Sohail2018} have found evidence of residual population stratification in effect size estimates from large-scale GWAS employing both linear and mixed model approaches, suggesting that existing structure corrections may be inadequate. For traits controlled by a few large effect loci this confounding effect is likely to be relatively small, but for analyses of complex traits with thousands of putatively associated SNPs the potential for confounding is high. Here we simulate a range of spatially correlated phenotypes for simulated individuals and seek to identify conditions under which existing GWAS structure corrections are subject to bias caused by continuous population structure.

@@

%We compared our model output to a two-dimensional stepping-stone model and a one-population coalescent model implemented in msprime \citep{Kelleher2016}. To isolate spatial effects from other components of the model such as overlapping generations and density-dependent probability of survival, we also ran a second version of the SLiM model with random mate selection and dispersal.

We ran 400 simulations for the spatial and random-mating SLiM models on a square landscape of size 50 with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), average lifetime $L=4$, genome size $1\times10^{8}$, recombination rate $1\times10^{-9}$, and drawing $/sigma$ values from a uniform distribution bounded by 0.2 and 4. To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM and later applied mutations to the coalesced tree sequence with msprime's 'mutate( )' function. Because msprime assumes that timesteps are in units of generations, we divided the per-generation rate of $1\times10^{-8} mut/site/gen$ by the generation time estimated for each value of $\sigma$ (see 'Genealogical Parameters' below) to convert the rate to units of mutations/site/timestep. We then verified that this procedure produced the correct number of mutations by comparing a subset of simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime (Supplementary Figure 1). Simulations were run for 1.6 million timesteps (approximately 30N generations), or until all extant individuals shared a common ancestor within the simulation (i.e. the tree sequence had coalesced).

%Coalescent stepping-stone simulations were on a 5x5 grid with 10 haploid samples per deme and migration between adjacent populations. To facilitate comparison across simulation types we ran one coalescent simulation for every SLiM parameter set. Because $\sigma$ is approximately equal to the mean distance between parent and offspring, we approximate the relationship between $\sigma$ and the migration rate in a grid simulation as $\sigma=4md$, where $m$ is the migration rate and $d$ is the ratio of landscape widths in continuous space vs coalescent simulations (here, 10). That is, an individual in an interior population has a $4m$ probability of being a migrant, and if it is a migrant will move $d$ units on the map. We set the total $N_{e}$ equal to the census size of each SLiM simulation divided by the variance in reproductive output (see below).

\subsection{Genealogical Parameters}
To study the geographic spread of genealogical ancestry in our simulations we output the full pedigree of all individuals over 50 generations with $\sigma$ set to 0.2, 0.35, 0.5, 1, 2, and 3.5 (corresponding to Wright's Neighborhood Sizes of 2.5, 7.7, 15.7, 62.8, 251.3, and 769.7, respectively). We then selected the individual closest to the center of the landscape in the final generation of each simulation, pruned the pedigree to include only its genealogical ancestors (that is, regardless of the amount of genetic material actually inherited from a given ancestor) and plotted the age of its most recent ancestor in each region of the full landscape using the 2-dimensional summary feature of ggplot2 in R \citep{Wickham2016}. We calculated the rate of spread of genealogical ancestry by estimating the mean and maximum distances between the focal individual and its ancestors in each generation and then fitting a linear model for distance as a function of generations using ordinary least squares regression.

We also estimated the variance in offspring and generation time for all simulations. To estimate generation times we stored the age of all parents for 200 timesteps and took the mean. To estimate variance in offspring we output the complete pedigree of all individuals in SLiM, counted the number of offspring for all individuals born in timesteps 50-150 (to ensure that the simulation captured the full lifespan of the sampled individuals), and calculated the variance.

\subsection{Sampling}
Our model records the genealogy and sequence variation of the complete population, but in real data genotypes are only observed from a a relatively small number of sampled individuals. We modeled three sampling strategies similar to common data collection methods in empirical genetic studies (Figure 1). "Random" sampling selects individuals at random from across the full landscape, "point" sampling selects individuals proportional to their distance from four equally spaced points on the landscape, and "midpoint" sampling selects individuals in proportion to their distance from the middle of the landscape. Downtream analyses were repeated across all sampling strategies.

\subsection{Summary Statistics}
We calculated the site frequency spectrum and a set of 18 summary statistics (Supplementary Table 1) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{is}$), and Tajima's D, as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, which we summarized as the correlation coefficient between $log10(spatial distance)$ and the proportion of identical base pairs among individuals.

@@

@@

\subsection{Genealogical Parameters and Run Times}
\subsection{Impacts of Sampling on Summary Statistics}
For spatial models the mean and variance of all summary statistics other than those involving the mean of the IBS block size distribution was significantly different across sampling strategies (Supplementary Table 1). Means were not significantly different across sampling strategies for any statistic under the random mating model. Only spatial correlation coefficients had significantly different variance under the random mating model, though in all cases the mean correlation was near zero.

@@

The inflation in $F_{is}$ appears to be caused by a deficit of heterozygous individuals in low-dispersal simulations -- a continuous-space version of the Wahlund effect \citep{Wahlund1928}. However under midpoint sampling observed heterozygosity is inflated over the random mating expectation, which we confirmed is caused by a higher proportion of heterozygotes in the middle of the landscape (Supplementary Figure S2 \cjb{CJ - this fig isn't super convincing -- will try denser sampling}). \cite{Shirk2014} observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.

@@

\subsubsection{Haplotype Block Lengths}
Trends in pairwise haplotype block sharing parallel those in SNP-based diversity estimates. At low dispersal the distribution of IBS block lengths in a set of samples is shifted towards smaller values -- resulting in lower means and fewer long IBS blocks. The variance and skew of the full IBS tract distribution vary across sampling strategies, but are similar to random-mating models under random sampling.

@@

\subsection{Demographic Modeling}
Demographic models from Stairwayplot tended to infer patterns of ancient population increases and recent declines when neighborhood sizes were low under all sampling strategies (Figure 3). Inflated past population sizes were common in both point and random sampling, demonstrating that the relatively minor shift in the site frequency spectrum observed with random sampling is enough to alter demographic estimates. Inference of severe population bottlenecks was also common across neighborhood sizes, particularly under midpoint and point sampling. Across all sampling schemes and neighborhood sizes around 20\% of runs inferred a sharp increase in population in the last 100 years.

Estimates stabilize for all sampling schemes as neighborhood sizes approach 750, and trend towards roughly half the census size. 
\cjb{I still need to figure out what's going on with Ne here... pi is $\approx$ expected value for a WF population with $N\approx N_{c}/2$, but variance Ne is more like 1/5 census N. I thought it was closer to 1/2 earlier because of a bug in my script where I was dropping individuals who had 0 offspring. Working on it...}.

@@

@@

@@

\section{Acknowledgements}
\newpage

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=\textwidth]{pop_params.pdf}
\caption{Genealogical parameters from spatial and random mating SLiM simulations, by neighborhood size.}
\label{fig:spectrum}
\end{figure*}

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{sfs_w_sumstats.pdf}
\caption{A: Site frequency spectra when sampling individuals from the midpoint, from four equally spaced points, and randomly across the landscape. Random mating models showed no differences by sampling scheme -- see supplementary figure S1. Spatial models have fewer low-frequency alleles and more mid-frequency alleles when neighborhood sizes are small, and this effect is stronger when sampling is more concentrated. B: Summary statistics under different sampling strategies, by neighborhood size.}
\label{fig:spectrum}
\end{figure*}

\afterpage{\clearpage}
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{stairwayplot_by_sigma_facet_sampling_by_bin.pdf}
\caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot by sampling scheme and neighborhood size (NS) range. Lines shown are the median inferred $N_{e}$ across 100 bootstrap replicates for each simulation. Colors are scaled to Wright's neighborhood size. The solid black line is the average census population size, and the dotted line is the average variance effective size ($N_{c}/var(n offspring)$).}
\label{fig:spectrum}
\end{figure*}

>>>>

%Space is the Place: How Dispersal and Competition Shape Genetic Variation in Continuous Space
%Impacts of Continuous Spatial Structure on Analyses of Population Genetic Data
%Impacts of Isolation by Distance on Demographic Modeling and Association Studies
%Space is the Place: How Dispersal and Sampling Shape Inference From Genetic Variation in Continuous Space
%Evolution on the Final Frontier: How Dispersal and Sampling Shape Inference From Genetic Data in Continuous Space
\title{Effects of Continuous Spatial Structure on Analysis of Population Genetic Data}

@@

\affil[$\ast$]{University of Oregon Dept. Biology, Institute for Ecology Evolution}
\keywords{Space; Population Structure; Demography; Haplotype block sharing; GWAS}

@@

\begin{abstract}
Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations. As a result many models used to infer selection, phenotypic associations, or past population sizes assume that samples are a random draw from a panmictic population. In reality most populations are structured to some degree across space, but the extent to which this model violation biases common genomic analyses is not well known. Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and association studies. We find that estimates of past population sizes and many common population genetic summary statistics vary significantly with both dispersal and sampling, with most bias occurring when neighborhood sizes are below 100 and sampling is spatially clustered. Variation in the site frequency spectrum and Tajima's D is greater across sampling strategies than over the range of demographic parameters we simulated, pointing to the need to carefully account for sampling when interpreting results. Last we show that the combination of isolation by distance and spatially correlated environments causes genome-wide signals of association with purely environmental phenotypes in linear-regression GWAS, and that this bias is not fully controlled by regressing out principal components positions during the analysis.  
\end{abstract}

@@

\section{Introduction}
The inescapable reality that biological organisms live, move, and reproduce in continuous spatial landscapes has been all but ignored in population genetic models. Indeed, a near universal rule of reproduction is that individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances. This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but classical mathematical models to describing are flawed approximations of the underlying process \citep{Felsenstein1975}.
As a result many studies focus on describing geographic structure as a mixture of random-mating populations connected by migration \citep{Malecot1948,Wright1931}, and analyze variation within clusters of genetic variation inferred by programs like \textit{structure} \citep{Pritchard2000} as random-mating units.

Though discrete models of between-population differentiation account for much of the genetic variation in species, the assumption that populations are randomly mating at some level has important implications for downstream inference of selection and demography. Methods based on the coalescent \citep{Kingman1982} assume that the sampled individuals are a random draw from a random mating population. If dispersal or mate selection is limited by space these assumptions are violated -- nearby individuals will be more closely related than an average random pair and drawing multiple samples from the same point on the landscape will not represent a random sample of the genetic variation present in the whole population. This is as true at the local scale as it is between the more differentiated units typically analyzed in phylogeographic studies. All downstream inferences drawn from the patterns of relationship inferred among sampled individuals will be subject to bias, but the extent depends on the degree of variation in ancestry across the landscape.

This issue is particularly important for analyses of polygenic selection, because the allele frequency variation caused by population structure is similar to the pattern expected when many sites in the genome are associated with a given phenotype \citep{Bulik-Sullivan2015}. Indeed two recent studies found that previous evidence of polygenic selection on human height in Europe was badly confounded by population structure \citep{Sohail2018,Berg2018}. As the scale of sequence data now available for many species allows inference of increasingly fine-scale patterns of selection and demography, understanding how and when subtle spatial structure is likely to bias results is an important task for population genetics.

Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness. The model scales to chromosome-scale alignments across tens of thousands of individuals, and outputs the full genealogy and recombination history of all final-generation individuals. We use this simulation to test how sampling strategy interacts with isolation by distance to cause systematic variation in population genetic summary statistics typically analyzed under discrete population models. We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum.

Last we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS). Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations \citep{Price2006}. Because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within continuous populations can also deflate $p$ values in GWAS of traits influenced by the environment \citep{Mathieson2012}. Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e. the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods of stratification bias corrections in GWAS are likely to fail.

\subsection{Modeling Evolution in Continuous Space}
The best-studied approaches to population genetics in continuous space were developed by \cite{Wright1943} and \cite{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. A key finding of Wright's model is that many important aspects of continuous populations can be described in terms of "neighborhood size" -- the number of potential mates for an individual in a given generation, defined as $4\pi\sigma^2d$, where $\sigma$ is the average dispersal distance and $d$ is population density. \cite{Maruyama1972} found that the rate of decline in genetic diversity in a 2-dimensional continuous population approaches the random mating expectation when $d\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

Though some aspects of continuous populations are fairly well described by the Wright and Malécot models, \cite{Felsenstein1975} showed that the assumptions of independent dispersal and Poisson distributed offspring that are the basis of these models are incompatible. Over time, a population meeting them will clump into a small number of geographic clusters occupying only a part of the available range. Although real populations are often clumped on landscapes due to factors like varying habitat quality and competition among species, the Wright and Malécot models produce much more extreme clumping than is observed in practice and fail to account for the density-dependent declines in population growth rate that are widely observed in real populations (CITE).

One method for modeling continuous populations is then to assume the existence of a grid of discrete randomly-mating populations connected by migration, which prevents clustering by forcing all regions to be occupied in every generation. Among many other important results drawn from this class of "lattice" or "stepping stone" models, \cite{Rousset1997} showed that the slope of the a linear regression of genetic differentiation ($F_{st}$) against the logarithm of spatial distance is an estimate of neighborhood size. Though good approximations of continuous structure given high dispersal, these models are not truly continuous, force a uniform realized population density across landscapes, and limit investigation of spatial structure below the level of the deme. An alternative method is to model the geographic spread of ancestry backwards in time through a diffusion approximation -- an approach that has recently made significant progress in modeling both dispersal and demographic parameters \citep{Barton2010,Kelleher2014,Ringbauer2017,Ringbauer2018}.

We took a direct approach to the clustering problem of classical forward-time models by incorporating density dependence. By scaling the probability of survival in each timestep to local population density we shift reproductive output towards regions of low-density and prevent populations from clustering. A similar approach was taken previously by \citep{Doebeli2003} who used an individual based model with continuous space and density dependent fitness to study the probability of speciation along continuous environmental gradients. However to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci as the unit of analysis, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

We ran 400 simulations for the spatial and random-mating SLiM models on a square landscape of size 50 with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), average lifetime $L=4$, genome size $1\times10^{8}$, recombination rate $1\times10^{-9}$, and drawing $/sigma$ values from a uniform distribution bounded by 0.2 and 4. To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM and later applied mutations to the coalesced tree sequence with msprime's 'mutate( )' function. Because msprime assumes that timesteps are in units of generations, we divided the per-generation rate of $1\times10^{-8} mut/site/gen$ by the generation time estimated for each value of $\sigma$ (see 'Genealogical Parameters' below) to convert the rate to units of mutations/site/timestep. We then verified that this procedure produced the correct number of mutations by comparing a subset of simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime. Simulations were run for 1.6 million timesteps (approximately 30N generations), or until all extant individuals shared a common ancestor within the simulation (i.e. the tree sequence had coalesced).

\subsection{Genealogical Parameters}
We recorded the census population size and estimated the variance in number of offspring and generation time for all simulations. To estimate generation times we stored the age of all parents for 200 timesteps after a 100 generation burn-in and took the mean. To estimate variance in offspring we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance across all individuals in timesteps 50-100.

\cjb{Would it be worthwhile to calculate the rate of spread of ancestry as a function of neighborhood size? IE how long until an individual in the middle of the map has ancestors in every region?}

\subsection{Sampling}
Our model records the genealogy and sequence variation of the complete population, but in real data genotypes are only observed from a relatively small number of sampled individuals. We modeled three sampling strategies similar to common data collection methods in empirical genetic studies (Figure 1). "Random" sampling selects individuals at random from across the full landscape, "point" sampling selects individuals proportional to their distance from four equally spaced points on the landscape, and "midpoint" sampling selects individuals in proportion to their distance from the middle of the landscape. Downstream analyses were repeated across all sampling strategies.

\subsection{Summary Statistics}
We calculated the site frequency spectrum and a set of 18 summary statistics (Supplementary Table X \cjb{working on this}) from 60 diploid individuals sampled from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. Statistics included common single-population summaries including mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{is}$), and Tajima's D, as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, which we summarized as the correlation coefficient between $log10(spatial distance)$ and the proportion of identical base pairs among individuals.

@@

In preliminary runs we found that inferred population histories were relatively variable even when simulating under a coalescent model, suggesting that some of the differences in demographic estimates for spatial models are caused by the behavior of the optimization algorithm rather than bias in the SFS caused by spatial mate choice and dispersal. To separate these effects we ran 100 coalescent simulations with constant population size $6.1\times 10^{-3}$ (the mean $N_{e}$ of random-mating SLiM models estimated from $pi$) and fit stairwayplot models using the same script as for our spatial models. We then calculated the standard deviation of inferred $N_{e}$ in each stairwayplot model to summarize the degree of fluctuation around the simulated population size, and asked if standard deviations were higher in spatial relative to coalescent models with a one-tailed t-test.

@@

\subsection{Genealogical Parameters and Run Times}
Census population size, generation time, and the variance in number of offspring all vary systematically with $\sigma$. Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000. Generation times vary between 5.2 and 4.9 timesteps per generation in the spatial model, and variance in offspring varies from 2.2 to 2.6.

%The fitness scaling procedure we use to avoid high fitness at range edges shrinks the suitable area by one $\sigma$ on all sides and likely explains part of the variation in census size, but scaling in offspring variance and the U-shaped distribution of census size in spatial models suggests additional processes at work. For spatial models census size initially rises as neighborhood sizes scale from 2 - 10, which seems to reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.

\subsection{Impacts of Sampling on Summary Statistics}
For spatial models the mean and variance of all summary statistics other than those involving the mean of the IBS block size distribution was significantly different across sampling strategies (Supplementary Table 1). Means were not significantly different across sampling strategies for any statistic under the random mating model. Only spatial correlation coefficients had significantly different variance under the random mating model, though in all cases the mean correlation was near zero.

@@

The inflation in $F_{is}$ appears to be caused by a deficit of heterozygous individuals in low-dispersal simulations -- a continuous-space version of the Wahlund effect \citep{Wahlund1928}. However under midpoint sampling observed heterozygosity is inflated over the random mating expectation, which we confirmed is caused by a higher proportion of heterozygotes in the middle of the landscape (Supplementary Figure S2). \cite{Shirk2014} observed a similar excess of heterozygosity in the middle of the landscape when simulating under a lattice model.

@@

\subsubsection{Haplotype Block Lengths}
Trends in pairwise haplotype block sharing parallel those in SNP-based diversity estimates (Supplementary Figure S1). At low dispersal the distribution of IBS block lengths in a set of samples is shifted towards smaller values -- resulting in lower means and fewer long IBS blocks. The variance and skew of the full IBS tract distribution vary across sampling strategies, but are similar to random-mating models under random sampling.

@@

\subsection{Demographic Modeling}
Demographic models from Stairwayplot tended to infer patterns of ancient population increases and recent declines when neighborhood sizes were below 20 under all sampling strategies (Figure 4). Inflated past population sizes were common in both point and random sampling, demonstrating that the relatively minor shift in the site frequency spectrum observed with random sampling is enough to alter demographic estimates. Inference of severe population bottlenecks was also common at neighborhood sizes under 100 for midpoint and point sampling strategies. Above neighborhood sizes of 100 the average inferred demography across all simulations was relatively accurate, with minor fluctuations slightly above the expected variance $Ne$. However individual model fits were highly variable and often inferred five-fold or greater population fluctuations even in high-dispersal simulations.

Standard deviations were significantly greater in all sampling strategies for neighborhood sizes under 20 and for midpoint sampling with neighborhood sizes 20-100, but all other sets performed similarly to the coalescent model. In summary, spatial mate choice and dispersal causes strong bias in SFS-based demographic estimates for neighborhood sizes below 20 or when sampling is clustered, but otherwise any biases are within the range of variability regularly inferred by Stairwayplot. In empirical data, it may be useful to run 10 or more replicate analyses on the same data and take a rolling average, which appears relatively robust for our simulations.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 13:50:47 2019
\begin{table}[ht]
\centering
\caption{\bf T-test results comparing standard deviations of inferred $N_{e}$ between spatial and coalescent models, by neighborhood size (NS) and sampling strategy. $p$ is the probability that spatial models have higher standard deviations.}
\begin{tabular}{rllrrrrr}
  \hline
sampling & NS range & t & df & $p$ \\ 
  \hline
random & 2-20 & 4.2572 & 41.6166 & 0.0001 \\ 
random & 20-100 & -1.8473 & 171.9905 & 0.9668 \\ 
random & 100-500 & -2.1297 & 164.3864  & 0.9827 \\ 
random & 500-1000 & -3.9681 & 147.0497 & 0.9999 \\ 
point & 2-20 & 7.0802 & 44.3615  & 0.0000 \\ 
point & 20-100 & -0.2038 & 169.3799 & 0.5806 \\ 
point & 100-500 & -2.4945 & 152.5000 & 0.9932 \\ 
point & 500-1000 & -3.8329 & 162.6443& 0.9999 \\ 
midpoint & 2-20 & 5.9253 & 59.5462 & 0.0000 \\ 
midpoint & 20-100 & 3.8940 & 171.7005  & 0.0001 \\ 
midpoint & 100-500 & -2.2764 & 139.5221 & 0.9878 \\ 
midpoint & 500-1000 & -3.2223 & 165.0792 & 0.9992 \\ 
   \hline
\end{tabular}
\end{table}

@@

\subsection{Sampling}

- clustered sampling causes more variation than dispersal scaling for several statistics
- range edges cause differences in the distribution of genetic variation across a range so sampling the midpoint gives a biased estimate of diversity and heterozygosity. Tajima's D is particularly sensitive to this effect. 
- in demographic analyses

\subsection{Demography}
- low dispersal and clustered sampling bias recent Ne down and inflate past Ne
- estimates are noisy in all simulations -- need to repeat analyses and average even for coalescent sims (???)
-

@@

- points: PCA correction is pretty good but not perfect. low-effect-size-alleles prob hit worst. 
- we know that this is also a problem empirically in human data, and was enough of an issue for arabidopsis that PC regression isn't used. Approaches like Bayesenv or the Science yellow warbler paper likely contain residual structure, particularly when environmenal clines parallel genetic clines and dispersal is high. 
- caveats: small populations, need to scale, mixed are probably likely better
- recs: run a gwas for sample x & y coordinates with whatever correction you're using. If there are significant values you're not correcting enough.
- for clinal traits, include x and y as covariates.

@@

\section{Acknowledgements}
\newpage
\newpage

@@

\begin{figure}[htbp]
\includegraphics{pop_params.pdf}
\caption{Genealogical parameters from spatial and random mating SLiM simulations, by neighborhood size.}
\label{fig:spectrum}
\end{figure}

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{sfs_w_sumstats.pdf}
\caption{A: Site frequency spectra under midpoint, point, and random sampling schemes. Lines are loess curves over the site frequency spectrum from each simulation with colors scaled to neighborhood size. Random mating models showed no differences by sampling scheme -- see supplementary figure S1. Spatial models have fewer low-frequency alleles and more mid-frequency alleles when neighborhood sizes are small, and this effect is stronger when sampling is more concentrated. B: Summary statistics under each sampling strategy, by neighborhood size.}
\label{fig:spectrum}
\end{figure*}

\afterpage{\clearpage}
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{stairwayplot_facet_rollmean.pdf}
\caption{Inferred demographic histories for spatial SLiM simulations from Stairwayplot, by sampling scheme and neighborhood size (NS) range. The thick line is a rolling mean and thin lines are individual model fits. Dashed horizontal lines are the average $N_{e}$ across random-mating SLiM models estimated from $\pi$.}
\label{fig:spectrum}
\end{figure*}
